A Bidirected String Graph Model for Genome Assembly

### 2012-12-12 presentation/paper.tex
fasta_head.pl
cleanup

\documentclass[10pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{mathptmx}
\usepackage{microtype}
\usepackage{array}
\usepackage{grffile}
\usepackage{threeparttable}
%\usepackage{hyperref}
%\usepackage{amsthm}

@@

\begin{table}[H]
	\begin{center}
	\caption{Approximate genome sizes of selected organisms}
		\label{tab:GenomeSizes}
		\begin{threeparttable}
			\begin{tabular}{|l|>{\arraybackslash}p{5.2cm}|}
				%\hline
				%$\phi$X174 bacteriophage &  $\sim$5,000 bp \\
				%\hline
				%Human mitochondrion &  $\sim$16,000 bp \\
				\hline
				{\bf Organism} & {\bf Approximate genome size} \\
				\hline
				{\it Variola} virus (smallpox) \tnote{1} &  186,000 bp \cite{smallpox}\\
				\hline
				{\it E. coli} (a bacterium) & 4,600,000 bp \\
				\hline
				{\it Ananas comosus} (domesticated pineapple) & 500,000,000 bp \\
				\hline
				{\it Zea mays} (domesticated corn) & 2,400,000,000 bp \cite{maize} \\
				\hline
				{\it Homo sapiens} (human)       &  3,200,000,000 bp \\
				\hline
			\end{tabular}
			\begin{tablenotes}
				\item [1] Technically, viruses are not organisms, but they still
				have genomes.  Actually, some viruses have genomes made of RNA,
				not DNA, but this is not considered in this paper.
			\end{tablenotes}
		\end{threeparttable}
	\end{center}
\end{table}

@@

Genomes may contain \KeyTerm{repeats}, which are subsequences that occur
multiple times in the genome--- possibly adjacent to each other (a
\KeyTerm{tandem repeat}), but not necessarily.  The number of times a
subsequence occurs in the genome is referred to as its \KeyTerm{copy number}.
Repeats may be longer than the length of any individual read, which makes the
genome more difficult to assemble accurately.  However, real genomes almost
always contain repeats of substantial length, so assemblers must be able to
handle repeats.

@@

Now that the idea of an exact overlap has been formalized, what we would like to
do is compute all pairwise exact overlaps of some minimum length $L$ among the
	reads that were provided as input to the assembler.  The chosen value of $L$
	must be balanced so that we find can enough overlaps to adequately cover the
	genome (small $L$ better) but still exclude most spurious overlaps (large
	$L$ better).

@@

The running time of this seed-and-extend algorithm is
typically much faster than the naive algorithm, although it is highly dependent
on the data.  The number of distinct seeds is at worst linear with respect to
the genome length.
%of a DNA sequence and its reverse-complement is the lexicographically lesser of
%the two.  Here, we refer canonical form of each seed, because we wish to
%compare reads that share a seed in either a forward or reverse-complement
%manner.}
For each of these distinct seeds, the reads containing it are examined pairwise,
which takes quadratic time with respect to the number of reads sharing the
seed.  Since more reads will contain each seed when the reads are sampled more
densely from the genome, the running time of the seed-and-extend algorithm rises
quadratically with respect to read density.  In addition, the number of read
pairs the algorithm considers for a specific seed also rises quadratically with
the number of times that seed is repeated in the genome due to genomic repeats.
Overall, the seed-and-extend algorithm is not ideal and is still a bottleneck of
the entire assembly algorithm, despite the fact that it is much faster than the
naive algorithm (except in degenerate cases, such as a ``genome'' that is
entirely a single repeated letter).

@@

After overlaps have been computed from the reads, the next step is to construct
a graph that models how the reads can be assembled when the overlaps are used as
evidence.  To do this, define a graph where each vertex corresponds to a
read\footnote{Actually, more than one read may map to the same vertex, as
explained in \Subsection{subsec:removing_contained_reads}.} and each edge
corresponds to an overlap.  Two vertices are connected by an edge if and only if
the corresponding reads share an overlap.  In addition, each edge is labeled
with an appropriate DNA sequence, the details of which will be explained later.

@@

	\begin{subfigure}{1.00\textwidth}
		\begin{center}
			\begin{tabular}{c|c}
				walk & interpretation \\ \hline
				$f \to g$ & $f.E \to g.E$ with sequence $g[o.g_{end} + 1 \dots ${\tt
				length}$(g)]$ \\
				$g \to f$ & $g.B \to f.B$ with sequence $f[o.g_{begin} - 1 \dots 0]$ \\
			\end{tabular}
		\end{center}
		\caption{Interpretation of each walk on the bidirected graph fragment}
	\end{subfigure}
	\caption{Full interpetation of an edge in a bidirected string graph}
	\label{fig:bidiedge_interp}
\end{figure}

@@

Building the graph is implemented by either the program
\ProgramName{build-bidirected-string-graph} or the program
\ProgramName{build-directed-string-graph}, the latter of which builds the
directed equivalent to the bidirected graph described in
\Subsection{subsec:directed}.  Both programs take as input two files: the set of
uncontained reads and their overlaps, and produce as output the bidirected or
directed graph, serialized to a file.
\ProgramName{build-bidirected-string-graph} will add one vertex for each read
and add a bidirected edge for each overlap, while
\ProgramName{build-directed-string-graph} will add two vertices for each read and
add two directed edges for each overlap.

@@

Collapsing unbranched paths is implemented as the program
\ProgramName{collapse-unbranched-paths}.  It takes as input a file containing a
string graph and produces as output a file containing the string graph with
unbranched paths collapsed.  Vertices are currently re-numbered; so, for
example, vertex 10 after collapsing unbranched paths does not necessarily refer
to the 10th read in the list of uncontained reads from which the graph was
constructed.

@@

\begin{figure}
	\begin{center}
		\vspace{1cm}
		\begin{tikzpicture}
			{\small
			\begin{tikzpicture}[>=triangle 45]
				\node[circle,fill=\nodecolor,draw=black] (A) at (0.0, 0) {A};
				\node[circle,fill=\nodecolor,draw=black] (B) at (3.0, 0) {B};
				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
			\end{tikzpicture} }
		\end{tikzpicture}
		\caption{
			Bidirected string graph for reads uniformly sampled from a 4.6 Mbp
			random genome after transitive reduction and path collapsing.  (Edge
			labeled with length of sequence, with the length being the same in both
			directions)}
		\label{fig:random_example}
	\end{center}
\end{figure}

@@

Note that in this file format, a continuous sequence can be split among multiple
lines.  A new sequence only begins after a line beginning with {\tt >} is
reached.

@@

where $n$ is the total number of reads, $G$ is the estimated genome length,
$\Delta$ is the length of the edge in bp, and $k$ is the number of reads that
support that edge.  To more precise, $k$ is the number of inner vertices in the
unbranched path corresponding to $e$ before the unbranched paths were collapsed
\cite{Myers2005}.

@@

$G$, the estimated genome length, is actually unknown, so it must be estimated
first.  It can be estimated by producing a ``bootstrap'' genome length by adding
up the lengths of the sequences on every edge in both directions and dividing by
2, then calculating the A-statistics using this bootstrap genome length and
identifying edges that are probably single-copy.  The A-statistics can then be
re-estimated over just these single-copy edges in order to refine the genome
length further before calculating the final A-statistic on all the edges.  This
algorithm is outlined in \cite{Myers2005}.

@@

Fortunately, there does exist a way to solve a bidirected flow problem by
mapping it to a directed flow problem by using the directed equivalent of the
bidirected graph, described in \Subsection{subsec:directed}.  However, the
calculated flow on a directed edge between, for example, two vertices $f.E
\to g.E$ is not guaranteed to be the same as the flow on $g.B \to f.B$, the edge
that corresponds to the other direction of the implied bidirected edge.  This
problem can be worked around by taking the average of the two counts and using
this as the flow on the bidirected edge.  This approximation is equivalent to
performing monotonization on the binet matrix that represents the linear program
reduction of bidirected flow to produce a totally unimodular matrix that
represents a linear program that can be solved to find an integal flow that
is at worst a 2-approximation to the optimal integral
flow on the bidirected graph\cite{Medvedev2009}\cite{Hochbaum2004}.

@@

\bibliographystyle{plain}
\bibliography{refs}

>>>>

\documentclass[12pt]{article}
\usepackage[left=3.3cm,right=3.3cm,bottom=3.0cm,top=3.0cm]{geometry}
\usepackage{float}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{subcaption}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{comment}
\usepackage{mathptmx}
\usepackage{microtype}
\usepackage{array}
\usepackage{grffile}
\usepackage{threeparttable}
%\usepackage{hyperref}
%\usepackage{amsthm}

\usepackage[nofiglist,notablist]{endfloat}
\renewcommand{\efloatseparator}{\mbox{}}

@@

\begin{table}[H]
	\begin{center}
	\caption{Approximate genome sizes of selected organisms}
		\label{tab:GenomeSizes}
		\begin{threeparttable}
			\begin{tabular}{|l|>{\arraybackslash}p{5.2cm}|}
				%\hline
				%$\phi$X174 bacteriophage &  $\sim$5,000 bp \\
				%\hline
				%Human mitochondrion &  $\sim$16,000 bp \\
				\hline
				{\bf Organism} & {\bf Approximate genome size} \\
				\hline
				{\it Variola} virus (smallpox) \tnote{1} &  186,000 bp \cite{smallpox}\\
				\hline
				{\it E. coli} (a bacterium) & 4,600,000 bp \\
				\hline
				{\it Ananas comosus} (domesticated pineapple) & 500,000,000 bp \\
				\hline
				{\it Zea mays} (domesticated corn) & 2,400,000,000 bp \cite{maize} \\
				\hline
				{\it Homo sapiens} (human)       &  3,200,000,000 bp \\
				\hline
			\end{tabular}
			\begin{tablenotes}
				\item [1] Technically, viruses are not organisms, but they still
				have genomes.
				%Actually, some viruses have genomes made of RNA,
				%not DNA, but this is not considered in this paper.
			\end{tablenotes}
		\end{threeparttable}
	\end{center}
\end{table}

@@

Genomes may contain \KeyTerm{repeats}, which are subsequences that occur
multiple times in the genome--- possibly adjacent to each other,
%(a \KeyTerm{tandem repeat})
but not necessarily.  The number of times a
subsequence occurs in the genome is referred to as its \KeyTerm{copy number}.
Repeats may be longer than the length of any individual read, which makes the
genome more difficult to assemble accurately.  However, real genomes almost
always contain repeats of substantial length, so assemblers must be able to
handle repeats.

@@

Now that the idea of an exact overlap has been formalized, what we would like to
do is compute all pairwise exact overlaps of some minimum length $L$ among the
	reads that were provided as input to the assembler.  The chosen value of $L$
	must be balanced so that we can find enough overlaps to adequately cover the
	genome (small $L$ better) but still exclude most spurious overlaps (large
	$L$ better).

@@

The running time of this seed-and-extend algorithm is
typically much faster than the naive algorithm, although it is highly dependent
on the data.  The number of distinct seeds is at worst linear with respect to
the genome length.
%of a DNA sequence and its reverse-complement is the lexicographically lesser of
%the two.  Here, we refer canonical form of each seed, because we wish to
%compare reads that share a seed in either a forward or reverse-complement
%manner.}
For each of these distinct seeds, the reads containing it are examined pairwise,
which takes quadratic time with respect to the number of reads sharing the
seed.  Since more reads will contain each seed when the reads are sampled more
densely from the genome, the running time of the seed-and-extend algorithm rises
quadraticly with respect to read density.  In addition, the number of read
pairs the algorithm considers for a specific seed also rises quadraticly with
the number of times that seed is repeated in the genome due to genomic repeats.
Overall, the seed-and-extend algorithm is not ideal and is still a bottleneck of
the entire assembly algorithm, despite the fact that it is much faster than the
naive algorithm (except in degenerate cases, such as a ``genome'' that is
entirely a single repeated letter).

@@

After overlaps have been computed from the reads, the next step is to construct
a graph that models how the reads can be assembled when the overlaps are used as
evidence.  To do this, define a graph where each vertex corresponds to a
read
%\footnote{Actually, more than one read may map to the same vertex, as explained
%in \Subsection{subsec:removing_contained_reads}.}
and each edge
corresponds to an overlap.  Two vertices are connected by an edge if and only if
the corresponding reads share an overlap.  In addition, each edge is labeled
with an appropriate DNA sequence, the details of which will be explained later.

@@

	\begin{subfigure}{1.00\textwidth}
		\begin{center}
			\begin{tabular}{c|c}
				walk & interpretation \\ \hline
				$f \to g$ & $f.E \to g.E$ with sequence $g[o.g_{end} + 1 \dots ${\tt
				length}$(g)]$ \\
				$g \to f$ & $g.B \to f.B$ with sequence $f[o.g_{begin} - 1 \dots 0]$ \\
			\end{tabular}
		\end{center}
		\caption{Interpretation of each walk on the bidirected graph fragment}
	\end{subfigure}
	\caption{Full interpretation of an edge in a bidirected string graph}
	\label{fig:bidiedge_interp}
\end{figure}

@@

Building the graph is implemented by either the program
\ProgramName{build-bidirected-string-graph} or the program
\ProgramName{build-directed-string-graph}, the latter of which builds the
directed equivalent to the bidirected graph described in
\Subsection{subsec:directed}.  Both programs take as input two files: the set of
uncontained reads and their overlaps, and produce as output the bidirected or
directed graph, serialized to a file.
\ProgramName{build-bidirected-string-graph} will add one vertex for each
uncontained read and a bidirected edge for each overlap, while
\ProgramName{build-directed-string-graph} will add two vertices for each
uncontained read and two directed edges for each overlap.

@@

Collapsing unbranched paths is implemented as the program
\ProgramName{collapse-unbranched-paths}.  It takes as input a file containing a
string graph and produces as output a file containing the string graph with
unbranched paths collapsed.  The algorithm marks vertices with in-degree 1 and
out-degree 1 as inner, then iterates through every non-inner vertex to locate
adjacent inner vertices, which must be the ends of unbranched paths.  Each
unbranched path is then walked and collapsed.  Vertices are currently
re-numbered; so, for example, vertex 10 after collapsing unbranched paths does
not necessarily refer to the 10th read in the list of uncontained reads from
which the graph was constructed.

@@

\begin{figure}
	\begin{center}
		\vspace{1cm}
		\begin{tikzpicture}
			{\small
			\begin{tikzpicture}[>=triangle 45]
				\node[circle,fill=\nodecolor,draw=black] (A) at (0.0, 0) {A};
				\node[circle,fill=\nodecolor,draw=black] (B) at (3.0, 0) {B};
				\draw[>->,very thick] (A) edge node[anchor=north] {4639100} (B);
			\end{tikzpicture} }
		\end{tikzpicture}
		\caption{
			Bidirected string graph for reads uniformly sampled from a 4.6 Mbp
			random genome after transitive reduction and unbranched path collapsing.  (Edge
			labeled with length of sequence, with the length being the same in both
			directions)}
		\label{fig:random_example}
	\end{center}
\end{figure}

@@

Note that in this file format, a continuous sequence can be split among multiple
lines; a new sequence only begins after a line beginning with {\tt >} is
reached.

@@

where $n$ is the total number of reads, $G$ is the estimated genome length,
$\Delta$ is the average length of $e$'s two sequences in bp, and $k$
%is the number of reads that support that edge.  To more precise, $k$
is the number of inner vertices in the unbranched path corresponding to $e$
before the unbranched paths were collapsed \cite{Myers2005}.

@@

$G$, the estimated genome length, is actually unknown, so it must be estimated
first.  It can be estimated by producing a ``bootstrap'' genome length by adding
up the lengths of the sequences on every edge in both directions and dividing by
2, then calculating the A-statistics using this bootstrap genome length and
identifying edges that are probably single-copy.  The estimated genome length
can then be refined using just these single-copy edges before calculating the
final A-statistic on all the edges.  This algorithm is outlined in
\cite{Myers2005}.

@@

Fortunately, there does exist a way to solve a bidirected flow problem by
mapping it to a directed flow problem by using the directed equivalent of the
bidirected graph, described in \Subsection{subsec:directed}.  However, the
calculated flow on a directed edge between, for example, two vertices $f.E \to
g.E$ is not guaranteed to be the same as the flow on $g.B \to f.B$, the edge
that corresponds to the other direction of the implied bidirected edge.  This
problem can be worked around by taking the average of the two counts and using
this as the flow on the bidirected edge.  This approximation is equivalent to
performing monotonization on the binet matrix that represents the linear program
reduction of bidirected flow; this produces a totally unimodular matrix that
represents a linear program that can be solved to find an integal flow that is
at worst a 2-approximation to the optimal integral flow on the bidirected
graph\cite{Medvedev2009}\cite{Hochbaum2004}.

@@

\addcontentsline{toc}{section}{References}
\bibliographystyle{plain}
\bibliography{refs}
